---
title: "**Linear Algebra in R**"
author: "Beau Thetford"
date: "19 May 2021"
output:
  html_document:
    df_print: paged
---

## Require Packages
```{r results="hide"}
require(matlib)
require(pracma)
require(ggplot2)
```

## Matrices
```{r}
A <- matrix(c(1,2,3,4), nrow = 2, byrow = TRUE)
B <- matrix(c(5,6,7,8), nrow = 2, byrow = TRUE)
A
B
```

## Math
```{r}
A+B #addition
A %*% B #matrix multiplication
A * B #entrywise multiplication
inv(A) #inverses

#pointing to entries
A[1,2] + B[2,2]

#slices
A[1,] #first row of A
B[,2] #second col of B

#dot product
x <- c(4,-3,5) #vectors
y <- c(5,6,8)

dot(x,y)

dot(A[,2], B[,2]) #dot columns of matrices

#trace (matlib)
tr(A)

#adjoint
adjoint(A)

#Frobenius inner product
tr(adjoint(A) %*% B)
```

# Systems of equations
Solve the system
\begin{align*}
  x_1 + 2x_2 &= 1 \\
  3x_1 -x_2 &= 4.
\end{align*}

```{r}
A <- matrix(c(1,2,3,-1), nrow=2 ) #coefficient matrix
A

b <- matrix(c(1,5), ncol=1) #rhs vector
b

aug <- cbind(A,b) #augmented 

rref(aug) #pacma rref function

x <- rref(aug)[,3] #isolate solution vector
x
```
# Functions

***Frobenius norm of a matrix***
```{r}
frob <- function(a){
  s = 0
  for (i in 1:nrow(a)){
    for (j in 1:ncol(a)){
      s = s + a[i,j]**2
    }
  }
  return(sqrt(s))
}

#test
x <- matrix(c(1,2,3,4), nrow=2)
print(frob(x))
```

## Matrices as linear transformations

***Rotation in the plane matrix*** 
```{r}
rotate <- function(x, theta){
  # accepts an ordered pair x and a rotation angle theta
  # returns x rotated by theta radians CCW
  
  R <- matrix(c(cos(theta), -sin(theta), sin(theta), cos(theta)), nrow = 2, byrow = T)
  return(R %*% x)
}

# test
x <- c(1,1)
rotate(x,pi/2)
```

***Polynomial differentiation matrix (power rule matrix)***
Let $a \in \mathbb{R}^{n+1}$ be the coefficient vector of a polynomial $f(x) \in P_n(\mathbb{R})$, with respect to the ordered basis $\beta = \{1, x, x^2, x^3, \dots, x^n\}$. Let $b \in \mathbb{R}^n$ be the coefficient vector of $\dfrac{df}{dx}$ in the basis $\gamma = \{1,x,x^2, \dots, x^{n-1}\}$. Then $b$ is given by 
$$ b = \left[\dfrac{df}{dx}\right]_\beta^\gamma a = 
\begin{pmatrix}0 & 1 & 0 & 0 & \cdots & 0 \\
0 & 0 & 2 & 0 & \cdots & 0 \\
0 & 0 & 0 & 3 & \cdots & 0\\
\vdots &\vdots &\vdots&\vdots&\ddots& \vdots \\
0 & 0 & 0 & 0 & 0& n\end{pmatrix}
\begin{pmatrix}a_0 \\ a_1 \\ a_2 \\ \vdots \\ a_n \end{pmatrix}, $$
where $\left[\dfrac{df}{dx}\right]_\beta^\gamma$ is the matrix of the derivative transformation $\dfrac{df}{dx} : P_n(\mathbb{R}) \rightarrow P_{n-1}(\mathbb{R})$ w.r.t the $\beta$ and $\gamma$ bases.

```{r}
polydiff <- function(a, n){
  # Accepts a vector (a) of polynomial coefficients relative
  # to the ordered basis {1, x, x^2, x^3, ...} (size n + 1, where (n) is
  # the degree of the polynomial).
  # Returns a vector of coefficients of the derivative relative
  # to the ordered basis for the subspace of dimension deg
  
  z <- matrix(data = 0, nrow = n, ncol = 1) # first column of zeros
  D <- matrix(data = 0, nrow = n, ncol = n) # allocate matrix
  for (i in 0:n){
    D[i,i] <- i # assign row index number to the diagonal entries
  }
  D <- cbind(z,D) # augment D to z column
  print('Differentiation Matrix:')
  print(D)
  print('df/dx coefficients:')
  return(D %*% a) # return differentiated coefficient vector
}

#test
x <- c(1, 0, 1) # 1 + 0x + 1x^2
polydiff(x, 2)
```




## Ill-conditioned systems
```{r}
condCheck <- function(A)
  if (cond(A) < 5 ){
    print("Great!")
  } else if (cond(A) > 10 & cond(A) < 50){
    print("Not so great")
  } else {
    print("ill-conditioned system!")
  }

condCheck(A)

cond(A)
# norm(inv(A)) * norm(A)
# fnorm(inv(A)) * fnorm(A)
```


## Eigenvectors and eigenvalues
```{r}
E <- eigen(A) #eigenvectors and values
V <- E$vectors #separate out vectors and values
lam <- E$values
```

Sanity check to make sure that $Av_1 = \lambda_1 v_1$, i.e. $Av_1 \cdot \dfrac{1}{\lambda_1} = v_1$:
```{r}
V[,1]
A %*% V[,1] / lam[1]
```





